Code used to generate paper results : Database comparison and CT characteristics

### 2024-05-16 Paper_code-CT_analysis.Rmd
Added all of Axelle's figures

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(tidyverse)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(msigdbr)
library(DOSE)
library(org.Hs.eg.db)
library(clusterProfiler)
library(patchwork)
library(BiocParallel)
library(Biostrings)
```

@@

load("../CTdata/eh_data/all_genes.rda")
load("../CTdata/eh_data/CCLE_data.rda")
load("../CTdata/eh_data/TCGA_TPM.rda")
load("../CTdata/eh_data/testis_sce.rda")
load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
load("../CTdata/eh_data/CT_genes.rda")

CT_and_CTP_genes <- CT_genes
CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
```

@@

```{r fig_param}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 6),
  title_gp = gpar(col = "black", fontsize = 6),
  row_names_gp = gpar(fontsize = 4),
  annotation_name_side = "left")

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
                   "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
                   "#9E0142")
chr_colors <- c("X-linked" = "deeppink", "Not X" = "royalblue1")

meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
```

# Pipeline figure values

@@

# CT genes characteristics

```{r basic}
table(CT_genes$testis_specificity)
table(CT_genes$transcript_biotype)
```

Most genes are testis specific 
(`r table(CT_genes$testis_specificity)[["testis_specific"]]/dim(CT_genes)[1]*100`%).
Most genes are mainly protein coding genes 
(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).

## X chromosome and regulated by methylation

In CTexploreR, genes have been characterised as regulated by methylation or not.

```{r X_meth}
table(CT_genes$X_linked)
table(CT_genes$regulated_by_methylation)
table(CT_genes$X_linked, CT_genes$testis_specificity)
table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
```

>>>>

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(tidyverse)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(msigdbr)
library(DOSE)
library(org.Hs.eg.db)
library(clusterProfiler)
library(patchwork)
library(BiocParallel)
library(Biostrings)
library(AUCell)
```

@@

load("../CTdata/eh_data/all_genes.rda")
load("../CTdata/eh_data/GTEX_data.rda")
load("../CTdata/eh_data/CCLE_data.rda")
load("../CTdata/eh_data/TCGA_TPM.rda")
load("../CTdata/eh_data/normal_tissues_multimapping_data.rda")
load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
load("../CTdata/eh_data/DAC_treated_cells_multimapping.rda")
load("../CTdata/eh_data/DAC_treated_cells.rda")
load("../CTdata/eh_data/testis_sce.rda")
load("../CTdata/eh_data/FGC_sce.rda")
load("../CTdata/eh_data/oocytes_sce.rda")

load("../CTdata/eh_data/CT_genes.rda")

CT_genes_types <- CT_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ "not_X_not_met")) %>% 
  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ "Testis specific",
                                        GTEX_category == "testis_preferential" ~ "Testis preferential",
                                        GTEX_category == "lowly_expressed" ~ "Undetectable in GTEX")) %>% 
  mutate(specificity = factor(specificity, 
                                     levels = c("Testis specific",  "Undetectable in GTEX", "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", "not_X_not_met")))

@@

```{r fig_param}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 4),
  title_gp = gpar(col = "black", fontsize = 5, fontface = "bold"),
  simple_anno_size = unit(0.2, "cm"),
  row_names_gp = gpar(fontsize = 3),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.2),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.05, "cm"),
  legend_height = unit(1, "cm"),
  use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10)

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
                   "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
                   "#9E0142")

TCGA_colors <- c(
    "BRCA" = "midnightblue", "COAD" = "darkorchid2",
    "ESCA" = "gold", "HNSC" = "deeppink2",
    "LUAD" = "seagreen", "LUSC" = "seagreen3",
    "SKCM" = "red3")

CCLE_colors <- c(
    "Lung" = "seagreen3", "Skin" = "red3",
    "Bile_Duct" = "mediumpurple1", "Bladder" = "mistyrose2",
    "Colorectal" = "plum", "Lymphoma" = "steelblue1",
    "Uterine" = "darkorange4", "Myeloma" = "turquoise3",
    "Kidney" = "thistle4",
    "Pancreatic" = "darkmagenta", "Brain" = "palegreen2",
    "Gastric" = "wheat3", "Breast" = "midnightblue",
    "Bone" = "sienna1", "Head_and_Neck" = "deeppink2",
    "Ovarian" = "tan3", "Sarcoma" = "lightcoral",
    "Leukemia" = "steelblue4", "Esophageal" = "khaki",
    "Neuroblastoma" = "olivedrab1")

DAC_colors <- c(
    "B2-1" = "olivedrab2", "HCT116" = "lightcoral",
    "HEK293T" = "seagreen3", "HMLER" = "mediumpurple1",
    "IMR5-75" = "deeppink2", "NCH1681" = "steelblue2",
    "NCH612" = "red3", "TS603" = "darkmagenta")

Regulation_colors <- c("Methylation" = "indianred2", 
                       "Not methylation"= "cyan4")

chr_met_colors = c("X_met" = "mediumorchid2",
                   "X_not_met" = "mediumseagreen",
                   "not_X_met"= "darkorange1",
                   "not_X_not_met" = "dodgerblue1")

Chr_colors = c("X-linked" = "deeppink", "not X" = "lightgreen")

Testis_specificity_colors = c("Testis specific" = "mediumpurple2",
                              "Testis preferential" = "goldenrod1",
                              "Undetectable in GTEX" = "gray")

testis_colors <- c(
    "SSC" = "floralwhite", "Spermatogonia" = "moccasin",
    "Early_spermatocyte" = "gold",
    "Late_spermatocyte" = "orange",
    "Round_spermatid" = "red2",
    "Elongated_spermatid" = "darkred", "Sperm" = "purple",
    "Sertoli" = "gray",
    "Leydig" = "cadetblue2", "Myoid" = "springgreen3",
    "Macrophage" = "gray10",
    "Endothelial" = "steelblue")

Fetal_cell_colors = c( "F_PGC" = "pink", 
                       "F_GC" = "pink4", 
                       "F_oogonia" = "palevioletred3",
                       "F_oocyte" = "mediumorchid4", 
                       "M_PGC" = "lightblue1", 
                       "M_GC" = "steelblue3",
                       "M_pre_spermatogonia" = "darkblue")

oocytes_colors <- c("Growing oocytes" = "paleturquoise3", 
                  # "Fully grown oocytes" ="lightgreen", 
                  # "Metaphase I" = "mediumvioletred", 
                  "Metaphase II" = "mediumvioletred")

sex_colors <- c("F" = "palevioletred1", "M"= "skyblue2")

stage_colors <- c( "pre-meiotic" = "moccasin", 
                   "meiotic"= "lightsalmon",
                   "post-meiotic" = "maroon4",
                   "somatic" = "gray")

# CT selection

## Pipeline figure values

@@

## GTEX-TCGA-CCLE

Illustrate the selection using heatmaps with GTEx data, TCGA and CCLE

Function to print GTEX expression with specific annotations.
```{r GTEX_function}
print_GTEX_heatmap <- function(genes = NULL, units = "log_TPM", 
                               row_fontsize = 2,
                               column_fontsize = 6, 
                               h_width = 3,
                               h_height = 10,
                               split_rows_by = "regulation") {

  suppressMessages({
    database <- GTEX_data
  })

  database <- database[rowData(database)$external_gene_name %in% genes, ]

  mat <- assay(database)
  rownames(mat) <- rowData(database)$external_gene_name

  if (units == "log_TPM") mat <- log1p(mat)

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(Testis_specificity)
    rowSplit <- gene_type$Testis_specificity
  }
  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }

  if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }

  Regulation <- as.matrix(gene_type$Regulation)
  rownames(Regulation) <- gene_type$gene
  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
                                     annotation_legend_param = legends_param,
                                     simple_anno_size = unit(0.2, "cm"),
                                     col = list(Regulation = Regulation_colors),
                                     annotation_name_gp = gpar(fontsize = 0))

  chr_met <- as.matrix(gene_type$chr_met)
  rownames(chr_met) <- gene_type$gene
  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
                                  annotation_legend_param = legends_param,
                                  simple_anno_size = unit(0.2, "cm"),
                                  col = list(chr_met = chr_met_colors),
                                  annotation_name_gp = gpar(fontsize = 0))

  Chr <- as.matrix(gene_type$Chr)
  rownames(Chr) <- gene_type$gene
  row_ha_chr <- rowAnnotation(Chr = Chr,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(Chr = Chr_colors),
                              annotation_name_gp = gpar(fontsize = 0))

  specificity <- as.matrix(gene_type$specificity)
  rownames(specificity) <- gene_type$gene
  row_ha_ts <- rowAnnotation(specificity = specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  h <- Heatmap(mat[gene_type$gene, ],
               name = "log(TPM +1)",
               column_title = "Normal tissues (GTEx)",
               column_title_gp = gpar(fontsize = 6, face = "bold"),
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_title_gp = gpar(fontsize = 4, face = "bold"),
               border = TRUE,
               border_gp = gpar(lwd = 0.5),
               col = heatmap_colors,
               cluster_rows = TRUE,
               cluster_columns = TRUE,
               show_row_dend = FALSE,
               show_column_dend = FALSE,
               row_names_gp = gpar(fontsize = row_fontsize),
               column_names_gp = gpar(fontsize = column_fontsize),
               column_names_side = "top",
               # right_annotation = c(row_ha_regulation, 
               #                        row_ha_chr, 
               #                        row_ha_ts,
               #                        gap = unit(1, "mm")),
               row_names_side = "left",
               width = unit(h_width, "cm"),
               height = unit(h_height, "cm"),
               clustering_method_rows = "ward.D",
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10,
               heatmap_legend_param = legends_param)

  return(h)
}
```

Function to print TCGA expression with specific annotations and to downsample 
the number of tumor samples per tumor type.

```{r TCGA_function}
print_TCGA_heatmap <- function(genes = NULL,
                               tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                            "TCGA-LUSC", "TCGA-COAD", 
                                            "TCGA-BRCA", "TCGA-HNSC",
                                            "TCGA-ESCA"),
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               h_width = 5, 
                               h_height = 10,
                               clustRow = TRUE,
                               clustCol = TRUE) {

  tcga <- TCGA_TPM
  tcga <- tcga[,tcga$shortLetterCode != "NT"]

  ## downsample tumors to have the same number of tumors in each type
  TCGA <- tcga[, tcga$project_id == tum_type[1]]
  TCGA <- TCGA[, 1:min(ncol(TCGA), downsample_to_n)]

  for (tum in tum_type[-1]){
    tmp <- tcga[, tcga$project_id == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    TCGA <- cbind(TCGA, tmp)
  }

  TCGA$project_id <- gsub(pattern = "TCGA-", x = TCGA$project_id, 
                          replacement = '')

  split_by <- factor(TCGA$project_id)

  column_ha_tumor <- HeatmapAnnotation(
    TCGA_tumor = TCGA$project_id,
    border = TRUE,
    col = list(TCGA_tumor = TCGA_colors),
    annotation_label = gpar(fontsize = 4),
    annotation_name_gp = gpar(fontsize = 0),
    simple_anno_size = unit(0.3, "cm"),
    annotation_legend_param = legends_param)

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- mat[CT_genes$external_gene_name,]
  mat <- log1p(mat)

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

  rowSplit <- NULL
  if(!is.null(split_rows_by)){
   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(Testis_specificity)
    rowSplit <- gene_type$testis_specificity
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }

  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }

  if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }
  }

  Regulation <- as.matrix(gene_type$Regulation)
  rownames(Regulation) <- gene_type$gene
  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
                                     annotation_legend_param = legends_param,
                                     simple_anno_size = unit(0.2, "cm"),
                                     col = list(Regulation = Regulation_colors),
                                     annotation_name_gp = gpar(fontsize = 0))

  chr_met <- as.matrix(gene_type$chr_met)
  rownames(chr_met) <- gene_type$gene
  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
                                  annotation_legend_param = legends_param,
                                  simple_anno_size = unit(0.2, "cm"),
                                  col = list(chr_met = chr_met_colors),
                                  annotation_name_gp = gpar(fontsize = 0))

  Chr <- as.matrix(gene_type$Chr)
  rownames(Chr) <- gene_type$gene
  row_ha_chr <- rowAnnotation(Chr = Chr,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(Chr = Chr_colors),
                              annotation_name_gp = gpar(fontsize = 0))

  specificity <- as.matrix(gene_type$specificity)
  rownames(specificity) <- gene_type$gene
  row_ha_ts <- rowAnnotation(specificity = specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

  # same range as in GTEX
  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  Heatmap(mat[gene_type$gene, ],
          name = "TCGA (logTPM)",
          column_title = "TCGA tumor samples",
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = FALSE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = 0),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          right_annotation = c(row_ha_ts,
                               #row_ha_regulation,
                               #row_ha_chr,
                               gap = unit(1, "mm")),
          row_split = rowSplit,
          cluster_row_slices = FALSE,
          width = unit(h_width, "cm"),
          height = unit(h_height, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}

Function to print CCLE expression with specific annotations. Allows to 
downsample the number of tumor samples pet tumor type to show

```{r CCLE_function}
print_CCLE_heatmap <- function(genes = NULL,
                               tum_type = c("Skin", "Lung", 
                                            "Colorectal", 
                                            "Breast", 
                                            "Head_and_Neck",
                                            "Esophageal"),
                               h_width = 5,
                               h_height = 10,
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               clustRow = TRUE,
                               clustCol = TRUE){

  ## downsample tumors to have the same number of tumors in each type
  ccle <- CCLE_data
  CCLE <- ccle[, ccle$type == tum_type[1]]
  CCLE <- CCLE[, 1:min(ncol(CCLE), downsample_to_n)]

  for (tum in tum_type[-1]){
    tmp <- ccle[, ccle$type == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    CCLE <- cbind(CCLE, tmp)
  }

  split_by <- factor(CCLE$type)

  column_ha_tumor <- HeatmapAnnotation(
    CCLE_cell_type = CCLE$type,
    border = TRUE,
    col = list(CCLE_cell_type = CCLE_colors),
    annotation_label = gpar(fontsize = 4),
    annotation_name_gp = gpar(fontsize = 0),
    simple_anno_size = unit(0.3, "cm"),
    annotation_legend_param = legends_param)

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- mat[CT_genes$external_gene_name,]
  mat <- log1p(mat)

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(testis_specificity)
    rowSplit <- gene_type$testis_specificity
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }
  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }

   if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }

  Regulation <- as.matrix(gene_type$Regulation)
  rownames(Regulation) <- gene_type$gene
  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
                                     annotation_legend_param = legends_param,
                                     simple_anno_size = unit(0.2, "cm"),
                                     col = list(Regulation = Regulation_colors),
                                     annotation_name_gp = gpar(fontsize = 0))

  chr_met <- as.matrix(gene_type$chr_met)
  rownames(chr_met) <- gene_type$gene
  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
                                  annotation_legend_param = legends_param,
                                  simple_anno_size = unit(0.2, "cm"),
                                  col = list(chr_met = chr_met_colors),
                                  annotation_name_gp = gpar(fontsize = 0))

  Chr <- as.matrix(gene_type$Chr)
  rownames(Chr) <- gene_type$gene
  row_ha_chr <- rowAnnotation(Chr = Chr,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(Chr = Chr_colors),
                              annotation_name_gp = gpar(fontsize = 0))

  specificity <- as.matrix(gene_type$specificity)
  rownames(specificity) <- gene_type$gene
  row_ha_ts <- rowAnnotation(specificity = specificity,
                                           annotation_legend_param = legends_param,
                                           simple_anno_size = unit(0.2, "cm"),
                                           col = list(specificity = 
                                                        Testis_specificity_colors),
                                           annotation_name_gp = gpar(fontsize = 0))

  # heatmap_colors <- colorRamp2(
  #   seq(0, quantile(rowMax(mat), 0.8), length = 11),                 
  #   legend_colors)

  # Same range as GTEX and TCGA
  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  Heatmap(mat[gene_type$gene, ],
          name = "CCLE (logTPM)",
          column_title = paste0("CCLE tumor cell lines"),
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = TRUE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          # right_annotation = c(row_ha_ts,
          #                      row_ha_regulation,
          #                      #row_ha_chr,
          #                      gap = unit(1, "mm")),
          row_split = rowSplit,
          width = unit(h_width, "cm"),
          height = unit(h_width, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}
```

Show CT genes in GTEX, CCLE and TCGA (downsampling the number of samples by 
tumor type).

```{r selection_heatmap}
my_genes <- CT_genes %>% 
  pull(external_gene_name)

ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              split_rows_by = "specificity")

ht_CCLE <- print_CCLE_heatmap(my_genes, 
                              downsample_to_n = 50, 
                              tum_type = c("Skin", "Lung", 
                                           #"colorectal", 
                                           #"breast", 
                                           #"esophageal",
                                           "Head_and_Neck"),
                              #"Leukemia", "Lymphoma"),
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "specificity")

ht_GTEX <- print_GTEX_heatmap(my_genes, 
                              row_fontsize = 1,
                              column_fontsize = 5,
                              h_width = 5,
                              h_height = 8,
                             split_rows_by = "specificity")
h_list <- ht_GTEX + ht_CCLE + ht_TCGA
draw(h_list, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
```

```{r, eval = FALSE, echo=FALSE}
pdf(file = paste0("figs/GTEX_Tumors.pdf"))
draw(h_list, main_heatmap = "TCGA (logTPM)", 
     ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
dev.off()
```

## Multimapping Fig

Figure showing how we kept the lowly_expressed in GTEx

```{r multimapping}
my_genes <- CT_genes %>% 
  filter(GTEX_category == 'lowly_expressed') %>% 
  filter(testis_specificity == "testis_specific")

without_MP <- log1p(
  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
    "TPM_no_multimapping"))
with_MP <-  log1p(
  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
    "TPM_with_multimapping"))
rownames(without_MP) <- rownames(with_MP) <- 
  rowData(normal_tissues_multimapping_data)[my_genes$ensembl_gene_id,]$external_gene_name

# Fix same range of colors for both heatmaps
full_mat <- cbind(without_MP, with_MP)
heatmap_colors <- colorRamp2(
  seq(0, quantile(rowMax(full_mat), 0.8), length = 11), legend_colors)

# Or same range of colors than GTEX heatmap
# heatmap_colors <- colorRamp2(
#   seq(0, 6, length = 11), legend_colors)

colnames(without_MP) <- gsub("_", x = str_to_sentence(colnames(without_MP)), ' ')
colnames(with_MP) <- gsub("_", x = str_to_sentence(colnames(with_MP)), ' ')
ordered_tissues <- 
  c("Testis", colnames(without_MP)[- which(colnames(without_MP) == "Testis")])
h_without_MP <- Heatmap(without_MP[my_genes$external_gene_name, ordered_tissues],
                        name = "no MP",
                        column_title = "Multimapped reads are excluded",
                        column_title_gp = gpar(fontsize = 10, face = "bold"),
                        border = TRUE,
                        border_gp = gpar(lwd = 0.5),
                        col = heatmap_colors,
                        cluster_rows = TRUE,
                        cluster_columns = FALSE,
                        show_row_dend = FALSE,
                        show_column_dend = FALSE,
                        row_names_gp = gpar(fontsize = 4),
                        column_names_gp = gpar(fontsize = 6),
                        column_names_side = "top",
                        row_names_side = "left",
                        heatmap_width = unit(8, "cm"),
                        heatmap_height = unit(12, "cm"),
                        clustering_method_rows = "ward.D",
                        use_raster = TRUE,
                        raster_device = "CairoPNG",
                        raster_quality = 10,
                        show_heatmap_legend = FALSE,
                        heatmap_legend_param = legends_param)

h_with_MP <- Heatmap(with_MP[my_genes$external_gene_name, ordered_tissues],
                     name = "log(TPM)",
                     column_title = "Multimapped reads are counted",
                     column_title_gp = gpar(fontsize = 10, face = "bold"),
                     border = TRUE,
                     border_gp = gpar(lwd = 0.5),
                     col = heatmap_colors,
                     cluster_rows = TRUE,
                     cluster_columns = FALSE,
                     show_row_dend = FALSE,
                     show_column_dend = FALSE,
                     row_names_gp = gpar(fontsize = 4),
                     column_names_gp = gpar(fontsize = 6),
                     column_names_side = "top",
                     row_names_side = "left",
                     heatmap_width = unit(8, "cm"),
                     heatmap_height = unit(12, "cm"),
                     clustering_method_rows = "ward.D",
                     use_raster = TRUE,
                     raster_device = "CairoPNG",
                     raster_quality = 10,
                     heatmap_legend_param = legends_param)

h_list <- h_without_MP + h_with_MP
draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
                 height = unit(5, "cm"))
```

```{r, eval = FALSE, echo = FALSE}
pdf(file = paste0("figs/Multimapping.pdf"))
draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
                 height = unit(5, "cm"))
dev.off()
```

## IGV

Shows why we removed illegitimate transcripts

```{r, echo=FALSE, fig.align='left', out.width = '100%', purl = TRUE}
knitr::include_graphics("figs/Fig_IGV_no_clear_transcript.png")
```

# CT_genes specific

From now on, we'll focus on CT genes that are specific

```{r}
CT_genes <- CT_genes %>%
  filter(CT_gene_type == "CT_gene") 
CT_genes_types <- CT_genes_types %>% 
  filter(CT_gene_type == "CT_gene")
```

# Methylation analysis

## Regulation by methylation

```{r meth_value_heatmap}
met_genes <- CT_genes %>% 
  filter(regulated_by_methylation) %>% 
  pull(external_gene_name)
met_value_number_per_gene <- tibble(
  gene = CT_genes$external_gene_name,
  regulated_by_methylation = CT_genes$regulated_by_methylation,
  NA_met_values_number = rowSums(
    is.na(assay(mean_methylation_in_tissues[CT_genes$external_gene_name, ]))))
n_met_values <- ncol(mean_methylation_in_tissues)
CT_met_with_WGBS <- met_value_number_per_gene %>%
  filter(regulated_by_methylation) %>% 
  filter(NA_met_values_number !=  n_met_values)

CT_met_without_WGBS <- met_value_number_per_gene %>%
  filter(regulated_by_methylation) %>% 
  filter(NA_met_values_number == n_met_values )

CT_not_met_with_WGBS <- met_value_number_per_gene %>%
  filter(!regulated_by_methylation) %>% 
  filter(NA_met_values_number !=  n_met_values)

CT_not_met_without_WGBS <- met_value_number_per_gene %>%
  filter(!regulated_by_methylation) %>% 
  filter(NA_met_values_number == n_met_values )

mean_methylation_fig <- function(genes = NULL, 
                                 values_only = FALSE,
                                 row_fontsize = 4,
                                 column_fontsize = 8, 
                                 na.omit = FALSE) {

  database <- mean_methylation_in_tissues
  rowData(database)$external_gene_name <- rownames(database)
  database <- database[genes, ]

  if (na.omit) {
    mat <- na.omit(assay(database))
    clustering_option <- TRUE
  } else {
    mat <- assay(database)
    clustering_option <- FALSE
  }

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes) %>% 
    mutate(group = case_when(gene %in% CT_met_with_WGBS$gene ~ "group1",
                             gene %in% CT_met_without_WGBS$gene ~ "group2",
                             gene %in% CT_not_met_with_WGBS$gene ~ "group3",
                             gene %in% CT_not_met_without_WGBS$gene ~ "group4")) %>% 
    left_join(CT_genes %>% dplyr::rename(gene = external_gene_name) %>% 
                dplyr::select(gene, somatic_met_level, sperm_met_level, 
                              germline_methylation, somatic_methylation)) %>% 
    arrange(group, sperm_met_level)

  # sort genes regulated by met differently than not regulated by met
  gene_type_met <- gene_type %>%
    filter(regulated_by_methylation)

  gene_type_not_met <- gene_type %>%
    filter(!regulated_by_methylation) %>%
    arrange(germline_methylation, somatic_met_level)

  gene_type <- rbind(gene_type_met, gene_type_not_met)
  colnames(mat) <- str_to_sentence(colnames(mat))

  h <- Heatmap(mat[gene_type$gene, sort(colnames(mat))],
               column_title = "Promoter methylation",
               name = "Mean \nmethylation",
               column_title_gp = gpar(fontsize = 6),
               border = TRUE,
               border_gp = gpar(lwd = 0.5),
               row_title_gp = gpar(col = c("indianred2", "cyan4"), fontsize = 6),
               row_split = factor(gene_type$Regulation,
                                  levels = c("Methylation", "Not methylation")),
               #row_split = factor(gene_type$group),
               cluster_row_slices = FALSE,
               col = colorRamp2(seq_len(100),
                                colorRampPalette(c("moccasin", "dodgerblue4"))(100)),
               na_col = "gray80",
               cluster_rows = FALSE,
               cluster_columns = FALSE,
               show_row_names = TRUE,
               show_heatmap_legend = TRUE,
               show_row_dend = FALSE,
               column_names_gp = gpar(fontsize = column_fontsize),
               row_names_gp = gpar(fontsize = row_fontsize),
               row_names_side = "left",
               heatmap_width = unit(4, "cm"),
               heatmap_height = unit(10, "cm"),
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 8,
               column_names_side = "top",
               heatmap_legend_param = legends_param)

  return(h)
}
```

```{r DAC_heatmap}
DAC_fig <- function(genes = gene_type$gene, multimapping = TRUE, 
                    row_fontsize = 2,
                    units = "log1p",
                    values_only = FALSE, scale_lims = NULL) {

  if (multimapping) {
    # database <- CTdata::DAC_treated_cells_multimapping()
    database <- DAC_treated_cells_multimapping

  } else {
    #database <- CTdata::DAC_treated_cells()
    database <- DAC_treated_cells
  }
